---
title: "Forecasting - Horizontal Solar Radition reaching the ground"
output:
  html_notebook: default
  html_document: default
---
</br>

### Student Details

* Saurabh Mallik 

</br>

<h3> Introduction </h3>

<b>TASK 1: </b></br>
The objective of task 1 is to analyse and forecast the horizontal solar radation reaching the ground at a particular location. We need to provide best 2 years forecasts using the following:
a. DLM Models (DLM, POLYDLM, KOYCK etc.)
b. Dynlm Models (SES, Holt's etc.)
c. ETS Models (AAA, MAA etc.)

The purpose of task 1 research is to understand and analyse which model in each of the three categories best fits the series and projects the forecasts.

The data to undertake this task are
1. DATA1.CSV (2 variables ie Solar Radiation and Precipitation and 660 observations) and 
2. Data.X.CSV (predictor series with 1 variable and 24 observations).

<b>TASK 2: </b></br>
The objective of this task is to analyse the correlation between quarterly residential property price index (PPI) and quartely population change over previous quarter in Victoria between September 2003 and December 2016. The dataset provided to carry out this investigation is Data2.csv.

Tha main aim of task 2 is to identify whether the correlation between the two is spurious or not.

</br></br>

<h3> Methodology </h3>

To undertake this research, forecasting methods on R Studio are being used to infer from the dataset.

</br>

<h3> Research and Inferences </h3>
<h5> 1. Task 1</h5>

```{r, message=FALSE, warning=FALSE, include=FALSE}
library(readr)
library(forecast)
library(TSA)
library(x12)
library(Hmisc)
library(AER)
library(tseries)
library(x13binary)
library(dLagM)
library(dynlm)
library(car)
library(xts)
```

Reading in the Data and preparing for analysis and converting to time series.
```{r}
data1 <- read_csv("~/Desktop/RMIT - Masters of Analytics/Year 1/Sem 2/MATH1307 - Forecasting/Forecasting - Ass 2/data1.csv")
datax <- read_csv("~/Desktop/RMIT - Masters of Analytics/Year 1/Sem 2/MATH1307 - Forecasting/Forecasting - Ass 2/data.x.csv")
datax.ts = ts(datax, start = c(2015,1), end = c(2016,12), frequency = 12)
solar = ts(data1$solar,start = c(1960,1), end = c(2014,12), frequency = 12)
ppt = ts(data1$ppt,start = c(1960,1), end = c(2014,12), frequency = 12)
data1.ts = ts(data1[,1:2])

```


Plotting the time series and checking for correlation.
```{r}
data.int = ts.intersect(solar , ppt)
colnames(data.int) =  c("Solar Radiation","Precipitation")
plot(data.int , yax.flip=T, main = "Fig1. Time series plot of solar radiation and Precipitation from 1960 to 2014")
```